---
title: "Are we more civilized today?"
author: "Adela Sobotkova"
date: "created: 10/4/2021, updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This exercise is based on the dataset provided by [OurWorldInData project](https://ourworldindata.org/homicides) based at the Oxford University.

## The long-term trend in Homicides in Western Europe

Understanding how homicide rates have changed prior to the modern era requires the help of historians and archivists. Manuel Eisner, a criminology professor at the University of Cambridge, and his colleagues published the Historical Violence Database \href{https://cjrc.osu.edu/research/interdisciplinary/hvd}: a compilation of data on long-term trends in homicide rates, in addition to qualitative information such as the cause of death, perpetrator and victim. This database is limited to countries with relatively complete historical records on violence and crime – mainly Western Europe and the US.

Starting in the second half of the nineteenth century, these European regions have consistent police records of those accused of murder or manslaughter and annual counts of homicide victims. To go back further in time, reaching as far back as the thirteenth century, Eisner collected estimates (from historical records of coroner reports, court trials, and the police) of homicide rates made in over ninety publications by scholars.5

In the chart shown here we see long-run homicide rates for the five European regions.

Homicide rates – measured as the number of homicides per 100,000 individuals – up to 1990 are sourced from Eisner’s (2003) publication and the Historical Violence Database. These estimates have been combined, from 1990 onwards, with homicide rates from the Global Burden of Disease study.6

In the 1300s, homicide rates in Western Europe ranged from 23 to 56 homicides per 100,000 individuals, comparable to the current rates in Latin America. Homicide rates then fell dramatically in all five regions from the 1300s to 1800s. The magnitude of this decline was similar across countries: homicide rates fell between 78 and 98%.

Homicide rates in Europe today are much lower than in the past. England, Germany, Switzerland, and Italy reached their lowest levels in the 2000s; while the Netherlands, Belgium, and Scandinavia reached their minimums in the mid to late 1900s.

```{r library}
library(tidyverse)
```

## Load the available data from ourworldindata.org

You should always interrogate the source of your data. Who compiled it, from where, what is missing, how representative the data are? Check the data/Metadata document to learn about the data provenance.

```{r data}
Western_Europe <- read_csv("data/homicide-rates-across-western-europe.csv")
```

## Inspect the data

How clean and analysis-ready is the dataset? Do you understand what the column names represent? What is the difference between rate and homicide number?

```{r inspect}
head(Western_Europe)
```



```{r}
Western_Europe <- Western_Europe %>% 
  rename(homicide_per_100k=`Homicide rate in Europe over long-term (per 100,000) (homicides per 100,000 people)`)
```

```{r}
Western_Europe <- Western_Europe %>% 
  rename(Country=Entity)
```


Note that the `echo = FALSE` parameter can be added to the code chunk to prevent printing of the R code that generated the plot.

Now, that you have looked at what the data looks like and what it represents, let's see what big picture it contains.

## Let's see what the long-term trend is in homicides
```{r }
ggplot(data = Western_Europe) + 
  geom_line(mapping = aes(x = Year, 
                           y = homicide_per_100k, 
                           color = Country)) +
  labs(x = "Year",
       y = "Number of Homicides per 100,000 people",
       title = "Homicide rate in Europe from 1300-2000")
```
Alright, the homicide rates are all descending over time. What a comfort. But the viz is not super clear. Let's check the rates for individual countries.

## Let's uncouple the homicides for easier view of individual countries

```{r }
ggplot(data = Western_Europe) + 
  geom_point(mapping = aes(x = Year, 
                           y = homicide_per_100k, 
                           color = Country)) +
  facet_wrap(~ Country, nrow = 3) +
  labs(x = "Year",
       y = "Homicides per 100,000",
       title = "Homicide rate in Western Europe from 1300-2000")

```

```{r}
# YOUR CODE HERE:
ggplot(data = Western_Europe) + 
  geom_line(mapping = aes(x = Year, 
                           y = homicide_per_100k, 
                           color = Country)) +
  facet_wrap(~ Country, nrow = 3) +
  labs(x = "Year",
       y = "Homicides per 100,000",
       title = "Homicide rate in Western Europe from 1300-2000") +
  theme(legend.position = "bottom")
```
## Let's cross-check the trends in homicide with the pattern of reign duration among Danish rulers through time

```{r}
kongeraekken <- read_csv2("data/kongeraekken_spreadsheet.csv")
```





```{r}
kongeraekken <- kongeraekken %>% 
  mutate(midyear=reign_last_year-(reign_last_year-reign_first_year)/2)
```

```{r}
ggplot(kongeraekken, aes(x = midyear, y = total_years_reign))+
  geom_point() +   #basic scatterpoint
  geom_smooth() +  #visual trend
  labs(title = "How long danish kings ruled over time",
       x= "year", y= "years they ruled") +   #better titles and axes's labels
  theme_bw()+                                #cleaner look
  theme(text = element_text(size = 14))      #bigger fond to make readable
```

## Answers to questions

Task 1: 
I renamed to long column name, so that the column is now called homicide_per_100k. See lines 51-54

Task 2: 
I chose to use geom_line, as I found that it made the visuability better. The lines make for a better view and understanding of development than the dots did. See line 95

Task 3: 
I did nothing because the layout already shows what has been asked as there are two columns and three rows. I tried to change the nrow to ncol - just to see if I thereby could make it opposite, and that worked. See line 98

Task 4: 
As in task 1 I renamed "Entity" to "Country" and afterwards I put it underneath the facets. See lines 56-59 and 102

Task 5: 
I added a floating table of content and made sure that the last day of updates would appear. This can be seen on the knitted version. See lines 4-8

Task 6: 
I uploaded my version of Kongeraekken; kongeraekken_spreadsheet, afterwards I made an extra column, named "midyear". I then made the graph with the titles x=year and y=year they ruled, and the two columns "midyear" on the x-axis and "total_years_reign" on the y-axis. See lines 104-127

Task 7: 
I think there is a problem when we try to judge civilazation by only one premise: homicide. There could be other several other factors that are relevant - just to mention some; violence or rape. When we want to make a statement on the development on our civilization, we need more information on different factors. Besides that, the numbers are not put in relation to the number of inhabitants, which could probably change the graphs a lot. However, there are of course clear signs to show that the amount of homicides are lowering, which tells us that at least on this parameter the population has become more civilized.




