---
title: "maternal_mortality"
author: "maria christensen"
date: "created: 12/3/2021, updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
---

libraries 
```{r}
library(tidyverse)
```

## Bringing data to the script
collecting the raw data on maternal mortality
```{r}
MMR <- read.csv("raw_data/maternal-mortality_raw_data_csv.csv",skip=1,sep = ";")
?read.csv
```

removing the previous data, that was not useful
```{r}
remove(maternal_mortality_csv)
remove(MMR_sel_contries)
remove(MMR_selected_contries)
```

collecting the raw data on births with skilled attendance 
```{r}
skilled_att <- read.csv("raw_data/births-attended-by-health-staff-sdgs_raw_data.csv", sep = ",")
```

## Tidying the spreadsheets
renaming columns to remove spaces and long and unintelligible names 
```{r}
MMR <- MMR %>% 
  rename(maternal_mortality_rate='Maternal.Mortality.Ratio..Gapminder..2010..and.World.Bank..2015..')

MMR <- MMR %>% 
  rename(country=Entity)

skilled_att <- skilled_att %>% 
  rename(Attended_births='Births.attended.by.skilled.health.staff....of.total.')

skilled_att <- skilled_att %>% 
  rename(Country=Entity)
```

## Collecting the 7 chosen countries
I am selecting the seven relevant countries in both maternal mortality rate and in skilled attendance at birth and giving them easily understandable names
```{r}
MMR_sel_countries <- MMR %>% 
  filter(country%in%c("Afghanistan","Chad","Italy","Nigeria","China", "Ethiopia", "India"))
```

```{r}
skilled_att_sel_countries <- skilled_att %>% 
  filter(Country%in%c("Afghanistan","Chad","Italy","Nigeria","China","Ethiopia","India"))
```


## Development in maternal mortality
```{r}
ggplot(data=MMR_sel_countries) +  
  geom_line(mapping = aes(x=Year, 
                          y=maternal_mortality_rate, 
                          color=country)) + 
  labs(x="Year", 
       y="Maternal mortality ratio", 
       title = "Maternal Mortality")
```

## Development in births with skilled attendance
```{r}
ggplot(data=skilled_att_sel_countries) +  
  geom_line(mapping = aes(x=Year, 
                          y=Attended_births, 
                          color=Country)) + 
  labs(x="Year", 
       y="Percent of births with skilled attendance", 
       title = "Skilled attended births")
```

## Afghanistan as an example of development
Looking at just one country: Afghanistan to see if there is a clear correlation
Afghanistan is one of the countries, where it is toughest to be pregnant, and therefore relevant to see the development here.
```{r}
MMR_afghanistan <- MMR %>% 
  filter(country=="Afghanistan")
```

```{r}
ggplot(data=MMR_afghanistan) +  
  geom_line(mapping = aes(x=Year, 
                          y=maternal_mortality_rate, 
                          color=country)) + 
  labs(x="Year", 
       y="Maternal mortality ratio", 
       title = "Maternal Mortality in Afghanistan")
```
```{r}
skilled_att_afghanistan <- skilled_att %>% 
  filter(Country=="Afghanistan")
```

```{r}
ggplot(data=skilled_att_afghanistan) +  
  geom_line(mapping = aes(x=Year, 
                          y=Attended_births, 
                          color=Country)) + 
  labs(x="Year", 
       y="Percent of births with skilled attendance", 
       title = "Skilled attended births in Afghanistan")
```

